% MedLock Gap Analysis — Section VI — Auto-generated

\subsection{Gap Analysis: MedLock vs. Existing Solutions}
\label{subsec:gap-analysis}

Existing clinical messaging systems exhibit critical gaps that MedLock's
architecture directly addresses. QuantumChat employs NTRU-HPS for
quantum resistance but lacks Zero-Trust enforcement—its perimeter-based
authentication permits lateral movement after initial access, violating
NIST SP 800-207 principles. EHR-PQC introduces ML-DSA (Dilithium) for
post-quantum signatures but operates within a classical TLS transport
layer, leaving key exchange vulnerable to harvest-now-decrypt-later
(HNDL) attacks. Neither system provides formal verification of its
security protocol.

Hypercare and HL7 FHIR achieve high throughput (750 and 620 RPS,
respectively) through classical ECDHE, but offer no post-quantum
protection and rely on implicit trust boundaries incompatible with
zero-trust mandates in HIPAA-regulated environments.

MedLock uniquely combines three capabilities absent from all competitors:
(1)~hybrid ML-KEM-768/X25519 key encapsulation providing both classical
and quantum-resistant confidentiality; (2)~full ZTA compliance with
per-request token validation, department-level micro-segmentation, and
cryptographic binding of producer identity; and (3)~formal ProVerif
verification proving that the protocol maintains secrecy and
authentication under a Dolev-Yao adversary model. While this hybrid
approach introduces a 2.0$\times$ latency overhead versus classical TLS
(605ms vs.\ 85ms p95), the 374 RPS throughput meets clinical SLA
requirements while providing provably quantum-safe communication—a
trade-off justified by the 15-year data retention mandates in healthcare.
